Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-18 paper/paper.Rmd
splits text to one line per sentence and adds some new comments

While human activities have altered the landscapes for millennials \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}. Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}. Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important \cite{With2019}.

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}. Related to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}. Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}. Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.

@@

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}. Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.

One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}. Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}. This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} with currently almost 300 packages in total. The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language. Thus, in the first part we are going to try to give a comprehensive overview over existing *R* packages for landscape ecology. In the second part, we will identify topics for which *R* packages are currently missing to the best of our knowledge. For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

<!-- MH: I went for a structure where in the first paragraph I briefly introduce the topic and then list the most common packages. -->

<!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->

<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->

@@

The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}. However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency and reproducibility. To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed. The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment.  Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. Popular approaches and packages include generalized linear models using e.g. the *stats* package; generalized additive models using e.g.the *mgcv* or *lme4* package; classification and regression trees (CART) using e.g. the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using e.g. the *ade4* or *vegan* package. Of course, also packages specifically designed for SDM exists. This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.

#### Connectivity {#sec:connectivity}
Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}. Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}. While structural connectivity focuses only on landscape characteristics (e.g. movement corridors, barriers, etc.), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.

@@

Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}. Such insights improve our understanding of metapopulation dynamics, speciation, species’ distributions and conservation \cite{Storfer2007}. By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than e.g. metapopulation genetics \cite{Holderegger2006}. As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.

@@

#### Neutral landscape models {#sec:NLM}
Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}. Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).

Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.

>>>>

While human activities have altered the landscapes for millenniums \cite{Ellis2011,Ellis2015}, in the past centuries the effects of humans on landscapes have increased to an unknown high, known as the the Anthropocene \cite{Crutzen2002}.
Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}. 
Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important \cite{With2019}.

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}. 
to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}. 
Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}. 
Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.

@@

<!-- JN:  We still can slightly update the landscape ecology introduction, but I think that the lenght of this text is good. Three paragraphs seems enough, as our goal is not to explain what is the landscape ecology, but what are the tools used in this field. -->

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. 
Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
<!--probably we should also mention the cost here-->
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.

One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}. 
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
<!-- we should add that the total number of spatial-related packages is, in fact, larger, as the task views, are manual lists, updated just by a few people -->
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming.
<!-- probably we should state why we are focusing on R only -->
<!-- some answers: "vastness of spatial R packages", ... -->
Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
In the second part, we identify topics for which *R* packages are currently missing to the best of our knowledge. 
For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

<!-- MH: I went for a structure where in the first paragraph I briefly introduce the topic and then list the most common packages. -->
<!-- JN: I like this idea -->

<!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->
<!-- JN: can you think of one/a few typical workflows? -->

<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write someting more about issues with archived packages in the second part (what's missing in R)... -->

@@

The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment. 
Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
<!-- JN: two notes: -->
<!-- 1. I am working on a paper supporting the belg package - I plan to submit it in less than a month, so hopefully we could update the reference here -->
<!-- 2. I also submitted a paper yesterday (2020-07-17) explaining a new package motif, which is aimed on calculations based on spatial signatures. I plan to add a paragraph about this package someplace in this paper. -->

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
Popular approaches and packages include generalized linear models using, e.g., the *stats* package; generalized additive models using, e.g., the *mgcv* or *lme4* package; classification and regression trees (CART) using, e.g., the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using , e.g., the *ade4* or *vegan* package. 
Of course, also packages specifically designed for SDM exists. 
This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
<!-- JN: I think all of the above packages should be cited -->

#### Connectivity {#sec:connectivity}

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}. 
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.

@@

Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}. 
Such insights improve our understanding of metapopulation dynamics, speciation, species’ distributions and conservation \cite{Storfer2007}. 
By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}. 
As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.

@@

#### Neutral landscape models {#sec:NLM}

Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).

Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. 
<!--JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields.
Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat.
-->
Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters. 
<!-- JN: I would also add landscapetools here as it allows to categorize outputs of the NLMR package -->

### 2020-07-18 paper/paper.tex
splits text to one line per sentence and adds some new comments

One example of a successful open-source project is the \emph{R}
programming language and its \emph{Comprehensive R Archive Network}
(CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming
language is among the most popular programming languages, especially in
ecology \cite{Lai2019}. Originally introduced as a statistical
programming language, a growing body of packages designed to analyze
spatial data subsequently emerged for the \emph{R} programming language
\cite{Bivand2006}. This is documented by the growing \emph{CRAN Task
View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task
View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020}
with currently almost 300 packages in total. The growing popularity of
the \emph{R} programming language for spatial data analysis and
landscape ecology can also be seen by the increasing number of related
textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist
\cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in
this review we are going to focus on tools implemented in the \emph{R}
programming language. Thus, in the first part we are going to try to
give a comprehensive overview over existing \emph{R} packages for
landscape ecology. In the second part, we will identify topics for which
\emph{R} packages are currently missing to the best of our knowledge.
For more general overviews see
e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

The use of landscape metrics was heavily facilitated by the introduction
of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published
in 1995 \cite{Kupfer2012,Gustafson2019}. However, \emph{FRAGSTATS} is
not not open-source software (however it is free to download at the
developers' homepage), which restricts transparency and reproducibility.
To provide a truly open-source solution to calculate landscape metrics,
recently the \emph{landscapemetrics} package \cite{Hesselbarth2019a} was
developed. The package allows to calculate the most widely used
landscape metrics in a transparent and reproducible workflow within the
\emph{R} environment. Additionally, the \emph{belg} packages allows to
calculate the Boltzmann entropy of a landscape gradient
\cite{Nowosad2019a}.

@@

Because the used modeling approaches are diverse
\cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of
\emph{R} packages used for SDMs. Popular approaches and packages include
generalized linear models using e.g.~the \emph{stats} package;
generalized additive models using e.g.the \emph{mgcv} or \emph{lme4}
package; classification and regression trees (CART) using e.g.~the
\emph{rpart}, \emph{randomForest} or \emph{ranger} package or
multivariate data analysis using e.g.~the \emph{ade4} or \emph{vegan}
package. Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat},
\emph{biomod2} and \emph{PresenceAbsence} packages.

@@

Connectivity is one of the core elements of landscape structure
\cite{Taylor1993} and thus one of the core concepts of landscape ecology
\cite{With2019}. Landscape connectivity describes how landscape
characteristics facilitate or hinder the movement of species
\cite{Tischendorf2000} or other aspects of mobility, such as dispersal,
gene or nutrient flow \cite{With2019}. While structural connectivity
focuses only on landscape characteristics (e.g.~movement corridors,
barriers, etc.), functional connectivity also includes behavior
characteristics of the species \cite{Tischendorf2000,With2019}.

@@

Landscape genetics investigates how characteristics of landscapes
interact with gene flow, genetic drift and selection \cite{Manel2003}.
Such insights improve our understanding of metapopulation dynamics,
speciation, species' distributions and conservation \cite{Storfer2007}.
By explicitly including characteristics of landscape, landscape genetics
is a way more realistic way of analysis than e.g.~metapopulation
genetics \cite{Holderegger2006}. As a result of its interdisciplinary,
landscape genetics combines methods from different disciplines including
landscape ecology, spatial statistics, geography and population genetics
\cite{Storfer2007}.

>>>>

One example of a successful open-source project is the \emph{R}
programming language and its \emph{Comprehensive R Archive Network}
(CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming
language is among the most popular programming languages, especially in
ecology \cite{Lai2019}. Originally introduced as a statistical
programming language, a growing body of packages designed to analyze
spatial data subsequently emerged for the \emph{R} programming language
\cite{Bivand2006}. This is documented by the growing \emph{CRAN Task
View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task
View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020}
that currently list almost 300 packages in total. The growing popularity
of the \emph{R} programming language for spatial data analysis and
landscape ecology can also be seen by the increasing number of related
textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist
\cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in
this review we are going to focus on tools implemented in the \emph{R}
programming. Thus, in the first part we are give a comprehensive
overview over existing \emph{R} packages for landscape ecology. In the
second part, we identify topics for which \emph{R} packages are
currently missing to the best of our knowledge. For more general
overviews see
e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

The use of landscape metrics was heavily facilitated by the introduction
of the \emph{FRAGSTATS} software \cite{McGarigal2012}, firstly published
in 1995 \cite{Kupfer2012,Gustafson2019}. However, \emph{FRAGSTATS} is
not not open-source software (however it is free to download at the
developers' homepage), which restricts transparency, reproducibility,
and extensibility. To provide a truly open-source solution to calculate
landscape metrics, recently the \emph{landscapemetrics} package
\cite{Hesselbarth2019a} was developed. The package allows to calculate
the most widely used landscape metrics in a transparent and reproducible
workflow within the \emph{R} environment. Additionally, the \emph{belg}
packages allows to calculate the Boltzmann entropy of a landscape
gradient \cite{Nowosad2019a}.

@@

Because the used modeling approaches are diverse
\cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of
\emph{R} packages used for SDMs. Popular approaches and packages include
generalized linear models using, e.g., the \emph{stats} package;
generalized additive models using, e.g., the \emph{mgcv} or \emph{lme4}
package; classification and regression trees (CART) using, e.g., the
\emph{rpart}, \emph{randomForest} or \emph{ranger} package or
multivariate data analysis using , e.g., the \emph{ade4} or \emph{vegan}
package. Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat},
\emph{biomod2} and \emph{PresenceAbsence} packages.

@@

Connectivity is one of the core elements of landscape structure
\cite{Taylor1993} and thus one of the core concepts of landscape ecology
\cite{With2019}. Landscape connectivity describes how landscape
characteristics facilitate or hinder the movement of species
\cite{Tischendorf2000} or other aspects of mobility, such as dispersal,
gene or nutrient flow \cite{With2019}. While structural connectivity
focuses only on landscape characteristics (e.g., movement corridors,
barriers), functional connectivity also includes behavior
characteristics of the species \cite{Tischendorf2000,With2019}.

@@

Landscape genetics investigates how characteristics of landscapes
interact with gene flow, genetic drift and selection \cite{Manel2003}.
Such insights improve our understanding of metapopulation dynamics,
speciation, species' distributions and conservation \cite{Storfer2007}.
By explicitly including characteristics of landscape, landscape genetics
is a way more realistic way of analysis than, e.g., metapopulation
genetics \cite{Holderegger2006}. As a result of its interdisciplinary,
landscape genetics combines methods from different disciplines including
landscape ecology, spatial statistics, geography and population genetics
\cite{Storfer2007}.
